%%%%%
%%
%% Sample document ``thesis.tex''
%%
%% Version: v0.2
%% Authors: Jean Martina, Rok Strnisa, Matej Urbas
%% Date: 30/07/2008
%%
%% Copyright (c) 2008-2011, Rok Strniša, Jean Martina, Matej Urbas
%% License: Simplified BSD License
%% License file: ./License
%% Original License URL: http://www.freebsd.org/copyright/freebsd-license.html
%%%%%

% !TEX TS-program = xelatex

% Available documentclass options:
%
%   <all `report` document class options, e.g.: `a5paper`>
%   withindex   - enables the index. New index entries can be added through `\index{my entry}`
%   glossary    - enables the glossary.
%   techreport  - typesets the thesis in the technical report format.
%   firstyr     - formats the document as a first-year report.
%   times       - uses the `Times` font.
%   backrefs    - add back references in the Bibliography section
%
% For more info see `README.md`
\documentclass[withindex]{cam-thesis}

% Citations using author-year format
\usepackage{natbib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Thesis meta-information
%%

%% The title of the thesis:
\title{Toward the Efficient, Scientific and Accessible Development of Small Language Models}

%% The full name of the author (e.g.: James Smith):
\author{Richard Diehl Martinez}

%% College affiliation:
\college{Churchill College}

\collegeshield{CollegeShields/Churchill}

%% Submission date [optional]:
% \submissiondate{November, 2042}

%% You can redefine the submission notice [optional]:
% \submissionnotice{A badass thesis submitted on time for the Degree of PhD}

%% Declaration date:
\date{August, 2025}

%% PDF meta-info:
\subjectline{Computer Science}
\keywords{Language Modeling, Small Language Models, Efficient Language Modeling}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Abstract:
%%
\abstract{%

Language models continue to grow in size, yet our understanding of their inner workings and ability to train them efficiently, particularly smaller models, remains limited. Small (sub-1 billion parameter) models offer practical advantages, including reduced financial and environmental costs, and broader accessibility, motivating the need for more effective training methodologies. This thesis addresses the challenge of developing small language models through two complementary lenses: cognitive inspiration and analytical investigation.

First, drawing parallels with efficient human language acquisition, we explore cognitively inspired techniques for training small models. We investigate curriculum learning strategies informed by human language acquisition in data-constrained settings, and we introduce Syntactic Smoothing, a cognitively motivated method that enhances the representation of infrequent words by leveraging syntactic structure.

Second, we adopt an analytical perspective to study the training dynamics and bottlenecks of small models. By analyzing the layer-wise behavior of the Pythia model suite, we identify convergence challenges and saturation phenomena in small models. This analysis exposes a broader shortcoming in current language model development: the disconnect between training and analysis tools, which hinders a scientific, iterative approach to model improvement. To address this, we introduce Pico, a open-source, lightweight, modular development framework for small models that integrates training and fine-grained analysis of model learning dynamics. Comprising pico-train and pico-analyze, Pico enables a principled, experiment-driven methodology for developing small language models. 

Ultimately, this thesis contributes novel techniques and tools aimed at making the training of small language models both more efficient, scientific and accessible to a wider range of users.
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Acknowledgements:
%%
\acknowledgements{%
   My thesis was supported by the Gates Cambridge Trust (grant OPP1144 from the Bill \& Melinda Gates Foundation). Part of the work in this thesis was funded by a grant from the Accelerate Programme for Scientific Discovery, made possible by a donation from Schmidt Futures. \\ This work would not be possible without the support of my advisor, Paula Buttery, my other academic mentors, Lisa Beinborn and Andrew Caines, and all the members of the `Office'. \\ Thank you to my dad for being my role model, my mom for being my rock and all previous generations who paved the way. I also want to acknowledge all my ride or dies, each of whom has brought something unique and joyful to my academic and personal journey. You know who you are: 
   \texttwemoji{dog}, \texttwemoji{mango}, \texttwemoji{giraffe}, \texttwemoji{violin}, \texttwemoji{climbing}, \texttwemoji{clap}, \texttwemoji{blossom}, \texttwemoji{rocket}, \texttwemoji{no_mouth}, and above all \texttwemoji{dog face}.

   
   % \texttwemoji{clapping-hands}, \texttwemoji{person-running}, \texttwemoji{rocket-ship}, 

   %  \texttwemoji{climbing}, \texttwemoji{clapping-hands},\texttwemoji{person-running}, \texttwemoji{rocket-ship}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Contents:
%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page, abstract, declaration etc.:
%% -    the title page (is automatically omitted in the technical report mode).
\frontmatter{}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Thesis body:

\input{chapters/introduction/main.tex}


% PART 1 

% Background section
\input{chapters/background/main.tex}

\input{chapters/climb/main.tex}

\input{chapters/syntatic-smoothing/main.tex}

% PART 2 

\input{chapters/analysis_background/main.tex}

\input{chapters/tending-towards-stability/main.tex}

\input{chapters/pico/main.tex}

\input{chapters/conclusion/main.tex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% References:
%%
% If you include some work not referenced in the main text (e.g. using \nocite{}), consider changing "References" to "Bibliography".
%

% \renewcommand to change default "Bibliography" to "References"
\renewcommand{\bibname}{References}
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{References}
\bibliographystyle{plainnat}
\bibliography{thesis}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Appendix:
%%

\appendix


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Index:
%%
\printthesisindex

\end{document}